---
output: github_document
---


<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE, 
  comment = "#>", 
  fig.path = "figures/README-",
  fig.retina = 3,
  out.width = "100%"
)
```

# Some Guidance for Writing `alt text` for Data Plots

## Why?

Alt text provides a textual alternative to non-text content in HTML documents. It serves various purposes: 

- it is necessary to for accessibility guidelines. 
- ensures your data visualisations communicate to everybody.
- assistive technologies can convert alt text into other formats such as speech or Braille, providing a description for people using screen reading software.
- alt text is displayed in place of the figure if it fails to load, or is viewed in a text-based browser. 
- alt text can assist both general and image-specific search engines.

## Ingredients for writing alt text

The purpose of alt text is to describe what is in a plot for a person who cannot see it. Generally, this means very basic explanation of the elements of the plots. You wouldn't typically interpret the plot, especially if it is also accompanied by a caption. A caption assumes that a person can see the plot, and provides additional context about why it is included and what is the most important thing to learn. Thus, the role of alt text is to provide the details of the plot elements for someone unable ot see it, which would help to make the caption also more understandable. When a caption doesn't exist, then the alt text has to play both roles.

The basic ingredients for alt text are:

- What kind of visualization is it (line graph, scatter plot, bar chart...)?
- What variables are on the axes?
- What is the range of the data?
- What is the aspect ratio.
- How are variables mapped to elements such as points, lines, violins or boxplots?
- What are the visible patterns (points are clustered in the top left and bottom right, lots of variability in size of the boxes of the boxplots, violins show a bimodal pattern, ...) without interpreting them.

## Examples

The examples are organised to show what would be written for some common plots, in both `alt text` and a `caption`. 

### Scatterplot 

The plot is generated from this code. 

```{r}
#| label: resid-plot
#| fig-width: 4
#| fig-height: 4
#| out-width: 40%
library(ggplot2)
library(broom)
cars_lm <- lm(mpg ~ hp, data = mtcars)
cars_all <- augment(cars_lm)
ggplot(cars_all, aes(x=.fitted, y=.resid)) + 
  geom_point()
```

**Alt text**: Scatterplot of fitted values on the x-axis and residual values on the vertical axis. The x-axis has tick marks 10, 15, 20, 25, and the y-axis has tick marks -5, 0, 5. There is a slight U-shape patterns to the points with one point with lower x-value than others at -12.5, and a high y-value of 7.5.

**Caption**: Examining the fit of a linear model of miles per gallon on horsepower. Residuals are plotted vertically and fitted values horizontally. The pattern suggests some non-linear relationship is not captured by the model.

## Resources

This is a repo that explains how to make alt-text for data plots

- https://medium.com/nightingale/writing-alt-text-for-data-visualization-2a218ef43f81
- https://lizharedogs.github.io/RLadiesNYAltText/#6
- https://medium.com/nightingale/writing-alt-text-for-data-visualization-2a218ef43f81
- https://www.w3.org/WAI/standards-guidelines/wcag/
- https://arbor-analytics.com/post/2023-06-30-making-r-graphs-more-accessible-using-brailler/
- https://www.w3.org/WAI/WCAG22/Techniques/html/H69.html

## How to contribute

We welcome contributions! If you have a suggestion or addition please post this as an [issue](https://github.com/numbats/alt-text-for-data-plots/issues) on GitHub.

